*Prochlorococcus marinus* responses to light and oxygen

### 2024-03-03 Code/Manuscript_O2.Rmd
Added to DNA discussion re: NATL2A and DNA repair

Figure \@ref(fig:ProchlorococcusEnzymeKms) shows the measured or inferred K~M~ for [O~2~] for genes encoding enzymes [@omarAnnotationGenesEncoding2023] for  *P. marinus* strains from clades HLI, LLI, LLIII and LLIV. XXXX CITATION  to Table once added XXXX. MED4 increases expression of alternative oxidase ('ubiquinol oxidase (non electrogenic)')  to cope with changes in light [@bergResponsesPsbAHli2011], by dissipating electrons from the inter-system transport chain. The approximate K~M~ for [O~2~] of ~ 25 µM for ubiquinol oxidase (non electrogenic) (Figure \@ref(fig:ProchlorococcusEnzymeKms)) is comparable to the lower limit for growth of MED4 in our experiments (Figure \@ref(fig:MED4GAM)).  We suggest that dependence upon this enzyme excludes MED4 from low oxygen zones.  The genome scan shows SS120 and MIT9313 lack this gene (Figure \@ref(fig:ProchlorococcusEnzymeKms)), and therefore, lack this oxygen-dependent path to cope with changing excitation.  Conversely, a gene encoding (S)-2-hydroxy-acid oxidase is encoded in the MIT9313 genome (Figure \@ref(fig:ProchlorococcusEnzymeKms)).  (S)-2-hydroxy-acid oxidase catalyzes the reaction of 2-hydroxy acid with O~2~ to produce toxic H~2~O~2~ [@cunaneCrystalStructureAnalysis2005]. (S)-2-hydroxy-acid oxidase has an approximate K~M~ for [O~2~] of ~ 250 µM, and produces H~2~O~2~, so growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffered photoinhibition, resulting from the inactivation of PSII caused by the production of the reactive oxygen species, hydrogen peroxide. This photoinhibition is compounded by the limited inducible repair mechanism for PSII, due to the absence of FtsH 1 and 2 XXXXCHECK TERMINOLOGYXXXX expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  We hypothesize that under the conditions of our high light and 2.5 µM or 25 µM O~2~ experiments, the activity of the (S)-2-hydroxy-acid oxidase enzyme is suppressed. As a result, the catalyzed production of hydrogen peroxide is inhibited, leading to less PSII damage, allowing MIT9313 to avoid photoinhibition and circumvent its limitations on PSII repair to exploit higher light.  Figure \@ref(fig:ProchlorococcusEnzymeKms) also shows that *P. marinus* SS120 is the only tested ecotype to lack the pyridoxal 5'-phosphate synthase enzyme. The pyridoxal 5'-phosphate synthase enzyme is an important cofactor in the biosynthesis of vitamin B~6~ [@sangIdentificationPyridoxinePyridoxamine2007]. Vitamin B~6~ is a potential antioxidant and can effectively quench singlet oxygen [@bilskiVitaminB6Pyridoxine2000]. The absence of the pyridoxal 5'-phosphate synthase enzyme may explain why *P. marinus* SS120 does not grow as well as *P. marinus* MIT9313, when exposed to high light stress under 25 µM O~2~ and not at all under 2.5 µM O~2~.

@@

Figure \@ref(fig:DNARepairFig) shows genes encoding DNA repair for *P. marinus* strains. As expected, *P. marinus* MED4 possesses the largest, most complete suite of genes encoding DNA repair enzymes, followed by *P. marinus* MIT9313. Conversely, *P. marinus* SS120 demonstrates the smallest genomic capacity for DNA repair. *Prochlorococcus marinus* MED4 was the only strain to possess a gene encoding deoxyribodipyrimidine photolyase, which, in the presence of blue light, is responsible for repairing DNA damaged by UV light [@sancarStructureFunctionDNA2003]. *Prochlorococcus marinus* MED4 was also the only strain to possesses a gene encoding DNA ligase, which uses ATP as a cofactor for DNA repair. The absence of  genes encoding deoxyribodipyrimidine photolyase and DNA ligase (ATP) in *P. marinus* MIT9313 and *P. marinus* SS120 helps explain why these two strains cannot tolerate growth under full [O~2~] and high light, found at the ocean surface. Furthermore, the protective effect of lower [O~2~], allowing these strains to grow at higher light, may relate in part to suppression of DNA damage when generation of Reactive Oxygen Species is suppressed at lower [O~2~]. NATL2A, a clade LLI, has been found near the ocean surface during deep ocean mixing [@malmstromTemporalDynamicsProchlorococcus2010]. Malmstrom *et al*. attributes NATL2a tolerance to short exposures of high light to the presence of the genes encoding photolyase, a gene found in HL clades which, in the presence of blue light, is responsible for repairing DNA damaged by UV light [@sancarStructureFunctionDNA2003].  *Prochlorococcus* are highly susceptible to hydrogen peroxide (H~2~O~2~) toxicity as they lack genes which scavenge H~2~O~2~ molecules [@morrisFacilitationRobustGrowth2008]. The small cell size of *Prochlorococcus* allow the reactive oxygen species (ROS), H~2~O~2~, to cross the cell membrane [@omarDiffusionalInteractionsMarine2022]; however, accumulation of extracellular H~2~O~2~ remains toxic to *Prochlorococcus* [@morrisFacilitationRobustGrowth2008,  @morrisDependenceCyanobacteriumProchlorococcus2011].

>>>>

Figure \@ref(fig:ProchlorococcusEnzymeKms) shows the measured or inferred K~M~ for [O~2~] for genes encoding enzymes [@omarAnnotationGenesEncoding2023] for  *P. marinus* strains from clades HLI, LLI, LLIII and LLIV. XXXX CITATION  to Table once added XXXX. MED4 increases expression of alternative oxidase ('ubiquinol oxidase (non electrogenic)')  to cope with changes in light [@bergResponsesPsbAHli2011], by dissipating electrons from the inter-system transport chain. The approximate K~M~ for [O~2~] of ~ 25 µM for ubiquinol oxidase (non electrogenic) (Figure \@ref(fig:ProchlorococcusEnzymeKms)) is comparable to the lower limit for growth of MED4 in our experiments (Figure \@ref(fig:MED4GAM)).  We suggest that dependence upon this enzyme excludes MED4 from low oxygen zones.  The genome scan shows SS120 and MIT9313 lack this gene (Figure \@ref(fig:ProchlorococcusEnzymeKms)), and therefore, lack this oxygen-dependent path to cope with changing excitation.  Conversely, a gene encoding (S)-2-hydroxy-acid oxidase is encoded in the MIT9313 genome (Figure \@ref(fig:ProchlorococcusEnzymeKms)).  (S)-2-hydroxy-acid oxidase catalyzes the reaction of 2-hydroxy acid with O~2~ to produce toxic H~2~O~2~ [@cunaneCrystalStructureAnalysis2005]. (S)-2-hydroxy-acid oxidase has an approximate K~M~ for [O~2~] of ~ 250 µM, and produces H~2~O~2~, so growth at lower [O~2~] may protect MIT9313 from auto-intoxication from production of H~2~O~2~. We hypothesize that under 250 µM O~2~ and higher blue light, *P. marinus* MIT9313 suffered photoinhibition, resulting from the inactivation of PSII caused by the production of the reactive oxygen species, hydrogen peroxide. This photoinhibition is compounded by the limited inducible repair mechanism for PSII, due to the absence of FtsH 1 and 2 XXXXCHECK TERMINOLOGYXXXX expression in *P. marinus* MIT9313 [@bonisteelStrainSpecificDifferences2018].  We hypothesize that under the conditions of our high light and 2.5 µM or 25 µM O~2~ experiments, the activity of the (S)-2-hydroxy-acid oxidase enzyme is suppressed. As a result, the catalyzed production of hydrogen peroxide is inhibited, leading to less PSII damage, allowing MIT9313 to avoid photoinhibition and circumvent its limitations on PSII repair to exploit higher light.  Figure \@ref(fig:ProchlorococcusEnzymeKms) also shows that *P. marinus* SS120 is the only tested ecotype to lack the pyridoxal 5'-phosphate synthase enzyme. The pyridoxal 5'-phosphate synthase enzyme is an important cofactor in the biosynthesis of vitamin B~6~ [@sangIdentificationPyridoxinePyridoxamine2007]. Vitamin B~6~ is a potential antioxidant and can effectively quench singlet oxygen [@bilskiVitaminB6Pyridoxine2000]. The absence of the pyridoxal 5'-phosphate synthase enzyme may explain why *P. marinus* SS120 does not grow as well as *P. marinus* MIT9313, when exposed to high light stress under 25 µM O~2~ and not at all under 2.5 µM O~2~ (Figure \@ref(fig:SS120PAR)).

@@

Figure \@ref(fig:DNARepairFig) shows genes encoding DNA repair for *P. marinus* strains. As expected, *P. marinus* MED4 possesses the largest, most complete suite of genes encoding DNA repair enzymes, followed by *P. marinus* MIT9313. Conversely, *P. marinus* SS120 demonstrates the smallest genomic capacity for DNA repair. *Prochlorococcus marinus* MED4 and NATL2A were the only strains to possess a gene encoding deoxyribodipyrimidine photolyase, which, in the presence of blue light, is responsible for repairing DNA damaged by UV light [@sancarStructureFunctionDNA2003]. *Prochlorococcus marinus* MED4 was also the only strain to possesses a gene encoding DNA ligase, which uses ATP as a cofactor for DNA repair. The absence of genes encoding deoxyribodipyrimidine photolyase and DNA ligase (ATP) in *P. marinus* MIT9313 and *P. marinus* SS120 explain why these two strains cannot tolerate growth under full [O~2~] and high light, found at the ocean surface. Furthermore, the protective effect of lower [O~2~], allowing these strains to grow at higher light, may relate in part to suppression of DNA damage when generation of Reactive Oxygen Species is suppressed at lower [O~2~]. NATL2A, a clade LLI, has been found near the ocean surface during deep ocean mixing [@malmstromTemporalDynamicsProchlorococcus2010]. Malmstrom *et al*. attributes NATL2A tolerance to short exposures of high light to the presence of the genes encoding photolyase, a gene found in HL clades. The presence of deoxyribodipyrimidine photolyase and absence of DNA ligase (ATP) supports why NATL2A tolerates limited exposure to high light and why NATL2A is unable to fully repair damaged DNA.  *Prochlorococcus* are highly susceptible to hydrogen peroxide (H~2~O~2~) toxicity as they lack genes which scavenge H~2~O~2~ molecules [@morrisFacilitationRobustGrowth2008]. The small cell size of *Prochlorococcus* allow the reactive oxygen species (ROS), H~2~O~2~, to cross the cell membrane [@omarDiffusionalInteractionsMarine2022]; however, accumulation of extracellular H~2~O~2~ remains toxic to *Prochlorococcus* [@morrisFacilitationRobustGrowth2008;  @morrisDependenceCyanobacteriumProchlorococcus2011].
